---
title: "Inferential Statistics Course Project"
author: "j-ros"
date: ''
output:
  word_document: default
  pdf_document: default
  html_document: default
---

##Exponential distribution: simultion analysis

###Introduction

In this section we are going to explore the statistical properties of the 
exponential distribution via averages of 40 exponentials. We will do 1000
simulations and compare the results with the theoretical expected ones.

###Setup

The code below simulates 1000 instances of 40 exponential distributions with 
rate parameter 0.2. It also reports the mean and variance of the simulated 
distribution.

```{r}
lambda<-0.2
n<-1000
m<-40
simRes<-array(rexp(n*m,lambda),c(n,m))
simMean<-apply(simRes,1,mean)
simVar<-apply(simRes,1,var)
sm<-mean(simMean); sv<-mean(simVar)
```

We can see that the sample mean `r sm` approaches the theoretical result of 
1/lambda = 5 and that the sample variance `r sv` approaches the theoretical
values of (1/lambda)^2 = 25, as per the Law of Large Numbers.

###Analysis

Now that we have our simulated distribution and parameters computed, we are 
going to show that the distribution resulting from simulations approaches a 
Gaussian one as stated by the Central Limit Theorem.

```{r}
hist(simMean,main="Distribution of means of 40 exponentials",probability=T,
     xlab="Mean of 40 exponentials")
curve(dnorm(x,mean=mean(simMean),sd=sd(simMean)),add=T,lwd=4)
```

##Basic Inferential Data Analysis

###Introduction
In this section we are going to perform basic inferential data analysis on the 
ToothGrowth dataset. First we explore the variables present and convert the dose
to factor. Then plot the len variable depending on supp and dose.

```{r}
str(ToothGrowth)
summary(ToothGrowth)
ToothGrowth$dose<-as.factor(ToothGrowth$dose)
plot(len~supp,data=ToothGrowth,xlab="Delivery method", ylab="Tooth length", main="Tooth length by Delivery method")
plot(len~dose,data=ToothGrowth,xlab="Dose", ylab="Tooth length", main="Tooth length by Dose")
```

Looking at the graphs, we are interested in testing whether or not the Tooth length for Delivery method OJ is significatively larger than VC. We are performing an independent group t-test for the unequal variance case at 95\% conficence level.

```{r}
x<-subset(ToothGrowth,supp=="OJ",len)
y<-subset(ToothGrowth,supp=="VC",len)
t.test(x,y,alternative="greater")
```

Since the p-value is smaller than 0.05 we find that Delivery method OJ produces larger
tooth length at a significant level.

We can similarly test the differences in dose, particularly Dose 2mg with respect the the other 2 in terms of Tooth length..

```{r}
x<-subset(ToothGrowth,dose=="0.5",len)
y<-subset(ToothGrowth,dose=="1",len)
z<-subset(ToothGrowth,dose=="2",len)
t.test(z,x,alternative="greater")
t.test(z,y,alternative="greater")
```

In both cases we find that Dose 2mg is significantly better than Dose 1mg and 0.5mg
even if we correct for false discovery rate, due to performing 2 tests on the same
dataset.